Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-08-22 TACO-paper-v2/coffee.tex
Draft (not finished) of the approach generality

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
aa

>>>>

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can dramatically impact the full-application performance. The next, logical question is to ask whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a generic vendor compiler should (could) integrate some of them.

We begin observing that, strategically, our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This is because there are neither conceptual restrictions nor technical reasons for which the compiler could not be integrated in frameworks other than Firedrake. It would be quite easy, for example, to achieve that with the popular FEniCS framework, the pioneer of automated code generation for finite element methods. However, more challenging is to assess the generality of the approach; that is, how COFFEE and its optimizations could be transferred to other computational domains, and whether that is really significant. To answer this question we need to jump back to the origins of the compiler.

In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization. Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest. The starting point of our work was a mathematical formulation of a local assembly operation, which is expressible as follows

\begin{equation*}
\scriptsize
\forall_{i, j} (A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) 
\end{equation*}

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$, computing the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ depend on the specific problem being solved and can be intricately complex, involving for example the evaluation of derivatives. Translating into code such a formulation of local assembly is a challenging process for complicated equations; for this reasons, research in automated code generation has been performed over the last few years. Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
\begin{enumerate}
\item All quantities are known values (apart obviously from the unknown $A$). This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no need to leave them as parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE. 
\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure lies at the core of generalized loop-invariant code motion. Whenever sub-expressions invariant with respect to different sets of loops can be found, the question of whether, where and how to hoist them arises. Any codes presenting these characteristics is amenable to this kind of transformation. 
\end{enumerate}

%In general, there can be a number of loops, with
